---
title: "Missing Data Sensitivity Analysis for Alcohol Research"
author: "Craig Enders"
date: "2026-04-26"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    theme: cosmo
    highlight: haddock
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '4'
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{css, echo=FALSE}
body {
  font-size: 18px;
}
pre {
  font-size: 15.6px;
  white-space: pre;
  overflow-x: visible;
}
pre code {
  white-space: pre;
  word-wrap: normal;
}
.main-container {
  max-width: 1500px !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  fig.width = 10,
  fig.height = 6)

options(scipen = 999)
```

This supplement presents `rblimp` R code and annotated output for models described in "Missing Data Sensitivity Analyses for Alcohol Research." The OSF repository at [https://osf.io/tp3k6](https://osf.io/tp3k6) includes the paper and supporting documents. The GitHub repository at
[https://github.com/craigenders/mnar-alc](https://github.com/craigenders/mnar-alc)
includes executable R and Blimp Studio scripts that apply the models from the paper to an artifical data set with the same features as the original.

The model numbers used throughout this supplement correspond to the models displayed in Tables 1, 3, and 4 in the paper. The seven sensitivity models are:

| Model | Description |
|:------|:------------|
| **Model 1** | CMAR |
| **Model 2** | CMAR With Auxiliary Variable |
| **Model 3** | Shared Parameter Model With Random Effects Predicting Dropout |
| **Model 5** | Shared Parameter Model With Random Effects and Their Squares Predicting Dropout |
| **Model 6** | Selection Model With Outcome Predicting Dropout |
| **Model 7** | Selection Model With Outcome and its Square Predicting Dropout |
| **Model 8** | Selection Model With Outcome and its Lag Predicting Dropout |

------------------------------------------------------------------------

## rblimp Arguments

The table below summarizes the major `rblimp` function arguments used
throughout this supplement. Each argument corresponds to a command in
the Blimp Studio scripts provided on the project's public GitHub site
([https://github.com/craigenders/mnar-alc](https://github.com/craigenders/mnar-alc)).
Readers unfamiliar with `rblimp` may find it helpful to use this table
as a reference while working through the analysis examples.

| Argument     | Description                                                            |
|:-------------|:-----------------------------------------------------------------------|
| `data`       | R data frame containing all variables                                  |
| `ordinal`    | Declares a variable as ordinal or categorical                          |
| `clusterid`  | Level-2 identifier variable                                            |
| `timeid`     | Integer variable indexing repeated measurement occasions               |
| `dropout`    | Creates the missingness or dropout indicator                           |
| `latent`     | Defines level-2 latent variables (random intercepts and slopes)        |
| `fixed`      | Completely observed predictor variables (speeds computation)           |
| `model`      | Specifies the focal outcome and dropout model regression equations     |
| `parameters` | Defines custom derived quantities computed from model parameters       |
| `nimps`      | Number of imputations (required for computing model-predicted probabilities for graphing) |
| `seed`       | Integer random number seed for reproducibility                         |
| `chains`     | Number of independent MCMC chains (4 is also the default)             |
| `burn`       | Number of warm-up iterations per chain (discarded before estimation)   |
| `iter`       | Number of post-burn-in iterations used to compute parameter summaries  |

------------------------------------------------------------------------

## Reading the Output

The table below describes each section of the `rblimp` output in the
order it appears. Detailed descriptions of the BURN-IN RHAT and OUTCOME
MODEL ESTIMATES sections follow the table.

| Output Section                      | Description                                                                                                                    |
|:------------------------------------|:-------------------------------------------------------------------------------------------------------------------------------|
| `BURN-IN POTENTIAL SCALE REDUCTION (PSR) OUTPUT`   | Assesses whether the MCMC warm-up period is sufficiently long; see detailed description below                                  |
| `DATA INFORMATION`                  | Gives information about the sample size and missing data rates                                                                 |
| `MODEL INFORMATION`                 | Reports the number of parameters and summarizes the fitted regression equations                                                |
| `MODEL FIT`                         | Reports DIC and WAIC and custom Wald tests, if specified                                                                       |
| `IMPUTED DATA SUMMARIES`            | Reports summary statistics for every manifest and latent variable in the model                                                 |
| `CORRELATIONS AMONG RESIDUALS`      | Reports correlations between pairs of residuals for a model's outcome variables                                                |
| `OUTCOME MODEL ESTIMATES`           | Parameter estimates, standard errors, credible intervals, and diagnostics; see detailed description below                      |
| `GENERATED PARAMETERS`              | Displays auxiliary parameters that are a function of the estimated parameters (e.g., time-specific standardized mean differences) |

### Potential Scale Reduction Convergence Diagnostic

The potential scale reduction factor (PSRF) or Rhat convergence
diagnostic appears near the top of the output. The PSRF assesses
whether the MCMC warm-up period is sufficiently long for the algorithm
to converge, which occurs when independent chains produce similar
parameter estimates. Smaller values closer to the theoretical minimum
of 1 are better, and values below 1.05 (Asparouhov & Muthén, 2010) or
1.10 (Gelman et al., 2014) are generally considered adequate. The
"Highest PSR" column displays the highest (worst) value across all
model parameters. This value should be less than 1.05 (or 1.10) by the
end of the burn-in period (the final row of the table). If the highest
PSRF is too large, increase the number of burn-in iterations (specified
using the `burn` argument).

### Outcome Model Estimates

Each outcome variable has its own summary table containing raw and
standardized coefficients, variances and standard deviations, and
variance explained effect sizes. The Estimate column gives the point
estimates (posterior median), the StdDev column displays "Bayesian
standard errors" (posterior standard deviations), the 2.5% and 97.5%
columns display 95% credible intervals. The ChiSq and PValue columns
are frequentist significance tests (Wald chi-square values).

The tables for b0i (random intercept) and b1i (random slope) summarize
the focal model's level-2 regression equations and random coefficient
variation. The table for dpdd displays the level-1 (within-person)
residual variance.

The table for dropout displays the dropout model parameters. The
dropout model's pseudo-R2 (the proportion of variance explained on the
latent response metric) is often the clearest indicator of model
quality. Values around .70 or higher suggest that the predictors almost
perfectly classify dropouts from completers, a condition known as
near-separation. Very large effect sizes almost always signal that a
model lacks support from the data.

### Effective Sample Size Diagnostic

The rightmost column in the parameter summary tables labeled N_Eff is the effective sample size (ESS)
diagnostic. The ESS provides a gauge of whether the post-warm-up
parameter summaries have sufficient precision. Numerically, it reflects
the effective number of MCMC iterations (the "sample size") that
contribute to these summaries after accounting for autocorrelation in
the sampling process. Values above 100 are generally considered
adequate (Gelman et al., 2014, p. 287). Inadequate PSRF and ESS values
indicate that more iterations are needed.

------------------------------------------------------------------------

## Setup

### Load Packages

```{r packages}
library(ggplot2)
library(patchwork)
library(rblimp)
set_blimp('/applications/blimp/blimp-nightly')
```

### Read Data

```{r load-data}
filepath <- 'https://raw.githubusercontent.com/craigenders/mnar-alc/main/mnar.csv'

mnar <- read.csv(filepath, stringsAsFactors = T)
```

### Source Plotting Functions

```{r source-functions}
source('https://raw.githubusercontent.com/blimp-stats/blimp-book/main/misc/functions.R')
source('https://raw.githubusercontent.com/craigenders/mnar-mlm/main/mnar-plotting.R')
```

### Helper Functions

```{r helper-functions}
# function to extract key parameter estimates
extract_params <- function(object, method) {

  tab <- object@estimates

  rows <- c(
    "b0i ~ Intercept",
    "b1i ~ Intercept",
    "b0i ~ med",
    "b1i ~ med",
    "b0i residual variance",
    "b1i residual variance",
    "Cor( b0i, b1i )",
    "dpdd residual variance",
    "Parameter: d_month0",
    "Parameter: d_month1",
    "Parameter: d_month2",
    "Parameter: d_month3",
    "Parameter: d_month4",
    "dropout R2: Coefficients"
  )

  res <- do.call(rbind, lapply(rows, function(r) {
    if (r %in% rownames(tab)) {
      round(tab[r, c("Estimate", "StdDev"), drop = FALSE], 2)
    } else {
      data.frame(Estimate = NA_real_, StdDev = NA_real_, row.names = r)
    }
  }))

  rownames(res) <- c(
    "Icept (M = 0)",
    "Slope (M = 0)",
    "Icept Diff.",
    "Slope Diff.",
    "Var(Icept)",
    "Var(Slope)",
    "Cor(Icept, Slope)",
    "Var(Residual)",
    "Std. Diff. T0",
    "Std. Diff. T1",
    "Std. Diff. T2",
    "Std. Diff. T3",
    "Std. Diff. T4",
    "Pseudo-Rsq"
  )

  colnames(res) <- c(
    paste0("Est_", method),
    paste0("SE_", method)
  )

  res
}

# function to extract convergence diagnostics
extract_convergence <- function(object, method) {

  neff <- object@estimates[, "N_Eff"]
  neff <- neff[is.finite(neff)]

  psr_row <- as.numeric(object@psr[20, ])
  psr_row <- psr_row[is.finite(psr_row)]

  data.frame(
    Min_Neff = round(min(neff),    3),
    Max_Neff = round(max(neff),    3),
    Min_PSR  = round(min(psr_row), 3),
    Max_PSR  = round(max(psr_row), 3),
    row.names = method
  )
}
```

------------------------------------------------------------------------

# Model 1: CMAR

Model 1 fits a two-level growth curve for drinks per drinking day
(dpdd) under the conditionally missing at random (CMAR) assumption.
This model serves as the baseline against which the MNAR sensitivity
models are compared.

```{r model1-cmar, cache=TRUE}
model1_cmar <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000)

output(model1_cmar)

```

------------------------------------------------------------------------

# Model 2: Auxiliary

Model 2 extends the CMAR model by incorporating an auxiliary variable
equation that regresses a baseline covariate (ftnd) on the random
intercept and slope. Including auxiliary variables that are correlated
with the outcome can recover missing information and improve the
plausibility of the MAR assumption.

*Important features of the auxiliary variable model:*

- The auxiliary variable equation regresses ftnd on the random effects
  b0i and b1i, leveraging the correlation between ftnd and the
  trajectory to recover missing information.

```{r model2-aux, cache=TRUE}
model2_aux <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # auxiliary variable equation
    ftnd ~ b0i b1i;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000)

output(model2_aux)

```

------------------------------------------------------------------------

# Dropout Probabilities

## Linear

The model in this section regresses the dropout indicator on month,
treatment (med), and their interaction to assess whether dropout rates
change linearly across occasions and whether that trend differs by
treatment group.

*Important features of the dropout model:*

- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).

```{r time-linear, cache=TRUE}
time_linear <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  model = 'dropout ~ intercept@-3 month month*med | intercept@0',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(time_linear)
```

## Quadratic

The model in this section extends the linear trend by adding a
quadratic month term, allowing dropout rates to accelerate or decelerate
across occasions.

*Important features of the dropout model:*

- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Squared terms are specified using `^2`.

```{r time-quadratic, cache=TRUE}
time_quadratic <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  model = 'dropout ~ intercept@-3 month month^2 month*med month^2*med | intercept@0',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(time_quadratic)
```

## Dummy-Coded

The model in this section uses dummy codes for each time point to
estimate occasion-specific dropout probabilities freely, without
imposing a functional form on how rates change over time.

*Important features of the dropout model:*

- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.

```{r time-dummy, cache=TRUE}
time_dummy <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  model = 'dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med | intercept@0',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(time_dummy)
```

## Dropout Probability Plots

```{r dropout-plots}
ymin <- 0
ymax <- .5

plot_pmiss_obs <- plot_means(dropout ~ month | med,
    model = time_dummy,
    ylab = "Probability",
    title = "Observed Probabilities",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_linear <- plot_means(dropout.1.probability ~ month | med,
    model = time_linear,
    ylab = "Probability",
    title = "Linear Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_quadratic <- plot_means(dropout.1.probability ~ month | med,
    model = time_quadratic,
    ylab = "Probability",
    title = "Quadratic Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_dummy <- plot_means(dropout.1.probability ~ month | med,
    model = time_dummy,
    ylab = "Probability",
    title = "Dummy-Coded Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss <- (plot_pmiss_obs | plot_pmiss_linear) / (plot_pmiss_quadratic | plot_pmiss_dummy)
plot_pmiss
```

## Obs. vs. Predicted

```{r dropout-fit}
pmiss_obs       <- aggregate(dropout ~ month + med,                        data = time_dummy@average_imp,     mean)
pmiss_linear    <- aggregate(dropout.1.probability ~ month + med,          data = time_linear@average_imp,    mean)
pmiss_quadratic <- aggregate(dropout.1.probability ~ month + med,          data = time_quadratic@average_imp, mean)
pmiss_dummy     <- aggregate(dropout.1.probability ~ month + med,          data = time_dummy@average_imp,     mean)

rmse_pmiss_linear    <- sqrt(mean((pmiss_linear$dropout.1.probability    - pmiss_obs$dropout)^2))
rmse_pmiss_quadratic <- sqrt(mean((pmiss_quadratic$dropout.1.probability - pmiss_obs$dropout)^2))
rmse_pmiss_dummy     <- sqrt(mean((pmiss_dummy$dropout.1.probability     - pmiss_obs$dropout)^2))

miss_fit <- data.frame(
  RMSE   = round(c(rmse_pmiss_linear, rmse_pmiss_quadratic, rmse_pmiss_dummy), 3),
  Min    = round(c(
    min(pmiss_linear$dropout.1.probability    - pmiss_obs$dropout),
    min(pmiss_quadratic$dropout.1.probability - pmiss_obs$dropout),
    min(pmiss_dummy$dropout.1.probability     - pmiss_obs$dropout)), 3),
  Median = round(c(
    median(pmiss_linear$dropout.1.probability    - pmiss_obs$dropout),
    median(pmiss_quadratic$dropout.1.probability - pmiss_obs$dropout),
    median(pmiss_dummy$dropout.1.probability     - pmiss_obs$dropout)), 3),
  Mean   = round(c(
    mean(pmiss_linear$dropout.1.probability    - pmiss_obs$dropout),
    mean(pmiss_quadratic$dropout.1.probability - pmiss_obs$dropout),
    mean(pmiss_dummy$dropout.1.probability     - pmiss_obs$dropout)), 3),
  Max    = round(c(
    max(pmiss_linear$dropout.1.probability    - pmiss_obs$dropout),
    max(pmiss_quadratic$dropout.1.probability - pmiss_obs$dropout),
    max(pmiss_dummy$dropout.1.probability     - pmiss_obs$dropout)), 3),
  row.names = c("Linear", "Quadratic", "Dummy")
)
```

```{r dropout-fit-display, echo=FALSE}
knitr::kable(miss_fit, caption = "Dropout Model Fit")
```

------------------------------------------------------------------------

# Model 3: Shared Parameter

Model 3 is a shared parameter model that links the probability of
dropout to each person's random intercept (b0i) and slope (b1i), so
that individuals with higher or lower trajectories have systematically
different dropout rates. Because scores are complete at baseline, the
dropout model uses a definition variable to activate these effects at
post-baseline occasions.

*Important features of the dropout model:*

- A definition variable (`preknot`) creates a time-varying dummy code
  that is 0 at baseline and 1 thereafter, activating dropout predictors
  at post-baseline occasions.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.

```{r model3-sharedparam, cache=TRUE}
model3_sharedparam <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # dropout model
    preknot = ifelse(month == 0, 0, 1);
    dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med
      b0i*preknot b1i*preknot | intercept@0;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(model3_sharedparam)

```

------------------------------------------------------------------------

# Model 5: Shared Parameter

Model 5 extends the shared parameter model by adding quadratic terms
for the random intercept and slope in the dropout equation, allowing
for nonlinear dependence of dropout on trajectory characteristics.
Because scores are complete at baseline, the dropout model uses a
definition variable to activate these effects at post-baseline
occasions. This model uses `burn = 60000` and `iter = 60000` to achieve
adequate convergence.

*Important features of the dropout model:*

- A definition variable (`preknot`) creates a time-varying dummy code
  that is 0 at baseline and 1 thereafter, activating dropout predictors
  at post-baseline occasions.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.
- Squared terms are specified using `^2`.

```{r model5-sharedparam, cache=TRUE}
model5_sharedparam <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # dropout model
    preknot = ifelse(month == 0, 0, 1);
    dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med
      b0i*preknot b1i*preknot (b0i^2)*preknot (b1i^2)*preknot | intercept@0;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 60000,
  iter = 60000,
  nimps = 20)

output(model5_sharedparam)

```

------------------------------------------------------------------------

# Model 6: Selection Model

Model 6 is a Diggle-Kenward selection model that conditions the
probability of dropout on the current outcome value (dpdd). Because
scores are complete at baseline, the dropout model uses a definition
variable to activate this effect at post-baseline occasions.

*Important features of the dropout model:*

- A definition variable (`preknot`) creates a time-varying dummy code
  that is 0 at baseline and 1 thereafter, activating dropout predictors
  at post-baseline occasions.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.

```{r model6-selection, cache=TRUE}
model6_selection <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # dropout model
    preknot = ifelse(month == 0, 0, 1);
    dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med
      dpdd*preknot | intercept@0;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(model6_selection)

```

------------------------------------------------------------------------

# Model 7: Selection Model

Model 7 extends the Diggle-Kenward model by adding a quadratic term for
the current outcome, allowing for nonlinear dependence of dropout on
outcome values. Because scores are complete at baseline, the dropout
model uses a definition variable to activate these effects at
post-baseline occasions.

*Important features of the dropout model:*

- A definition variable (`preknot`) creates a time-varying dummy code
  that is 0 at baseline and 1 thereafter, activating dropout predictors
  at post-baseline occasions.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.
- Squared terms are specified using `^2`.

```{r model7-selection, cache=TRUE}
model7_selection <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # dropout model
    preknot = ifelse(month == 0, 0, 1);
    dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med
      dpdd*preknot (dpdd^2)*preknot | intercept@0;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(model7_selection)

```

------------------------------------------------------------------------

# Model 8: Selection Model

Model 8 extends the Diggle-Kenward model by adding the lagged outcome
value, conditioning dropout on both the current and prior occasion's
dpdd. Because scores are complete at baseline, the dropout model uses a
definition variable to activate these effects at post-baseline
occasions.

*Important features of the dropout model:*

- A definition variable (`preknot`) creates a time-varying dummy code
  that is 0 at baseline and 1 thereafter, activating dropout predictors
  at post-baseline occasions.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean expressions create dummy codes for each time point, returning
  1 if MONTH = `t` and 0 otherwise.
- Appending `.lag` to a variable's name introduces its lagged value
  into the model.

```{r model8-selection, cache=TRUE}
model8_selection <- rblimp(
  data = mnar,
  ordinal = 'med',
  clusterid = 'id',
  timeid = 'month',
  dropout = 'dropout = dpdd (monotone)',
  latent = 'id = b0i b1i',
  fixed = 'month med',
  model = '
    # level-2 equations
    b0i ~ intercept@b0 med@b2;
    b1i ~ intercept@b1 med@b3;
    b0i b1i ~~ b0i b1i;
    # level-1 equation
    dpdd ~ intercept@b0i month@b1i;
    # dropout model
    preknot = ifelse(month == 0, 0, 1);
    dropout ~ intercept@-3 (month==1) (month==2) (month==3) (month==4)
      (month==1)*med (month==2)*med (month==3)*med (month==4)*med
      dpdd*preknot dpdd.lag*preknot | intercept@0;',
  parameters = '
    # compute time-specific effect sizes
    d_month0 = ( (b0+b2) - (b0) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month1 = ( (b0+b2 + 1*(b1+b3)) - (b0 + 1*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month2 = ( (b0+b2 + 2*(b1+b3)) - (b0 + 2*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month3 = ( (b0+b2 + 3*(b1+b3)) - (b0 + 3*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);
    d_month4 = ( (b0+b2 + 4*(b1+b3)) - (b0 + 4*b1) ) / sqrt(dpdd.totalvar + b0i.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 20000,
  iter = 20000,
  nimps = 20)

output(model8_selection)

```

------------------------------------------------------------------------

# Summary Tables

## Model Key

Model numbers correspond to Tables 1, 3, and 4 in the paper. Models 2 and 4 from those tables are omitted here because they are intermediate specifications not highlighted in the paper's primary comparisons.

| Model | Description |
|:------|:------------|
| **Model 1** | CMAR |
| **Model 2** | CMAR With Auxiliary Variable |
| **Model 3** | Shared Parameter Model With Random Effects Predicting Dropout |
| **Model 5** | Shared Parameter Model With Random Effects and Their Squares Predicting Dropout |
| **Model 6** | Selection Model With Outcome Predicting Dropout |
| **Model 7** | Selection Model With Outcome and its Square Predicting Dropout |
| **Model 8** | Selection Model With Outcome and its Lag Predicting Dropout |

## Parameter Estimates

The table displays the estimate (posterior median) and Bayesian standard error (posterior standard deviation) for each model. Column headers beginning with `Est` contain the point estimates (posterior medians) and those beginning with `SE` contain the Bayesian standard errors (posterior standard deviations).

```{r extract-params, cache=TRUE, dependson=c("helper-functions","model1-cmar","model2-aux","model3-sharedparam","model5-sharedparam","model6-selection","model7-selection","model8-selection")}
table_summary <- cbind(
  extract_params(model1_cmar,        "Mod1"),
  extract_params(model2_aux,         "Mod2"),
  extract_params(model3_sharedparam, "Mod3"),
  extract_params(model5_sharedparam, "Mod5"),
  extract_params(model6_selection,   "Mod6"),
  extract_params(model7_selection,   "Mod7"),
  extract_params(model8_selection,   "Mod8")
)
```

```{r table-summary-display, echo=FALSE, dependson="extract-params"}
knitr::kable(table_summary, caption = "Parameter Estimates")
```

## Changes in Standard Error Units

Each cell reports the difference between Model $k$ and the CMAR baseline (Model 1), scaled by Model 1's Bayesian standard error. Values near zero indicate that the MNAR assumption has little practical impact on that parameter relative to sampling variability.

```{r table-change, cache=TRUE, dependson="extract-params"}
est_cmar <- table_summary[, "Est_Mod1"]
se_cmar  <- table_summary[, "SE_Mod1"]

compare_methods <- c("Mod2", "Mod3", "Mod5", "Mod6", "Mod7", "Mod8")

table_change <- sapply(compare_methods, function(m) {
  round((table_summary[, paste0("Est_", m)] - est_cmar) / se_cmar, 2)
})
table_change <- as.data.frame(table_change)
rownames(table_change) <- rownames(table_summary)
```

```{r table-change-display, echo=FALSE, dependson="table-change"}
knitr::kable(table_change, caption = "Changes vs. CMAR in Standard Error Units")
```

## Convergence Diagnostics

```{r extract-conv, cache=TRUE, dependson=c("model1-cmar","model2-aux","model3-sharedparam","model5-sharedparam","model6-selection","model7-selection","model8-selection")}
table_diag <- rbind(
  extract_convergence(model1_cmar,        "Model 1"),
  extract_convergence(model2_aux,         "Model 2"),
  extract_convergence(model3_sharedparam, "Model 3"),
  extract_convergence(model5_sharedparam, "Model 5"),
  extract_convergence(model6_selection,   "Model 6"),
  extract_convergence(model7_selection,   "Model 7"),
  extract_convergence(model8_selection,   "Model 8")
)

table_diag$Iterations <- c(
  nrow(model1_cmar@iterations),
  nrow(model2_aux@iterations),
  nrow(model3_sharedparam@iterations),
  nrow(model5_sharedparam@iterations),
  nrow(model6_selection@iterations),
  nrow(model7_selection@iterations),
  nrow(model8_selection@iterations)
)
```

```{r table-diag-display, echo=FALSE, dependson="extract-conv"}
knitr::kable(table_diag, caption = "Convergence Diagnostics")
```

------------------------------------------------------------------------
